---
title: "The education impacts of metropolitan ecosystem disruption: evidence from invasive ash borer"
author: "Alberto Garcia"
date: "March 19, 2023"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: yes
    toc: no
  html_document: default
  documentclass: article
bibliography: eab_chicago.bib
header-includes: 
  \usepackage{setspace}
  \usepackage{bbm}  
  \usepackage{graphicx}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{sectsty}
  \usepackage{float}
  \usepackage{color}
  \usepackage{outlines}
  \usepackage{multirow}
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(kableExtra)
library(tidyverse)
library(Metrics)
library(reshape2)

knitr::opts_chunk$set(echo = FALSE, warnings = FALSE, messages = FALSE, out.width="49%", fig.align = "center")

palette <- list("white" = "#FAFAFA",
                "light_grey" = "#d9d9d9",
                "dark" = "#0c2230",
                "red" = "#ed195a",
                "blue" = "#1c86ee",
                "green" = "#7CAE7A",
                "dark_green" = "#496F5D",
                "gold" = "#DAA520")
```
<!--
TO DO LIST


-->

# Abstract {-}
This paper shows that invasive species driven ecosystem degradation has ramifications for education outcomes in metropolitan areas. I leverage variation from the introduction of an invasive insect that exclusively targets ash trees, the emerald ash borer, to the Chicago Metropolitan area. Exploiting the staggered and idiosyncratic spread of the borer, I show how tree canopy cover and education outcomes were affected. Difference-in-differences methods indicate that confirmed infestations reduced canopy cover in affected areas by roughly 2\% on average. Further, schools within 2.5 kilometers of a confirmed infestation site saw declines in standardized test performance, with impacts felt most strongly by low-income students on the lower end of the academic performance spectrum. There is also weak evidence of decreased attendance rates resulting from infestation detection. This paper shows that invasive species can substantially impact ecosystem service provision and ultimately, education outcomes, adding to the damages known to be caused by invasive forest pests.

\clearpage

\doublespacing


# Introduction

<!-- 1-2 paragraphs motivation (about puzzle or problem) -->

Invasive species can have significant impacts on environmental quality and provision of ecosystem services. There is a growing recognition of the need for research on the economic impacts of ecosystem disruption [@ferraro_causal_2019; @fenichel_control_2014], however, causal evidence that examines how changes in biological features of the environment affect human behavior is scarce [@frank_social_2023]. This is particularly true for the relationship between economic outcomes and invasive species induced ecosystem changes. Economists have only recently started to study the pathways through which invasive species affect economic outcomes [@jones_invasive_2020].

Prior research has established environmental drivers of educational outcomes including the impacts of temperature [e.g., @park2020] and pollution [e.g., @marcotte2017] on learning. Vegetation and greenery has long been associated with improved educational outcomes [e.g., @kweon2017], however, much of this work is purely correlational, and there exists little to no causal evidence outside of laboratory settings. Trees have been shown to mitigate extreme temperatures and air pollution in addition to poviding psychological benefits [@turner-skoff2019]. Each of these represent mechanisms through which tree cover loss may negatively affect education outcomes. 

<!-- 1 paragraph research question -->

In this paper, I ask whether invasive species induced tree cover change affects education outcomes in a metropolitan setting. To answer this question, I take advantage of the idiosyncratic spread and detection of the ash borer within the Chicago Metropolitan area. The emerald ash borer represents one of the most destructive invasive species in the United States. A beetle that exclusively targets ash trees, the ash borer has decimated the primary street tree used in many US cities. This includes Chicago, where prior to the arrival of the ash borer, ash were the most common found non-invasive tree species in the area's streets and parks (Morton Arboretum 2020). After its introduction to Illinois, survey efforts were conducted from 2008 to 2015 to understand the spread of the ash borer throughout the region. This spread provides idiosyncratic variation in canopy cover change and in the associated ecosystem services through time across the region. 

<!-- 2 paragraphs findings -->

I use a difference-in-differences approach to recover the causal impacts of the ash borer on both tree cover and education outcomes. I find that ash borer infestation led to a 2 percent reduction in the canopy cover of affected areas. Further, infestation led to relatively poorer attendance rates and standardized test performance at schools within 2.5 kilometers of a confirmed infestation site. This was particularly true for low-income students, where effects were concentrated. While infestation detection led to a reduction in meeting standards and an increase in academic warnings amongst non-low income students, there was no impact on the non-low income group. 

Several checks lend creedence to the causal interpretation of these findings. I provide event study estimates showing the growing impact of ash borer on canopy cover change through time. While ash borer arrival leads to an initial spike in rates of tree cover loss, longer term canopy cover decline comes, in part, from reduced tree cover gain in affected areas. This is consistent with work showing that tree planting budgets and priorities changed following the arrival of the pest [e.g., @hauer2017]. Further, the canopy cover impacts are greatest in areas where ash make up a greater proportion of an area's tree population. I also demonstrate the dose response of education outcomes to changes in canopy cover using an instrumental variables approach and find that canopy cover gain leads to improved standardized test performance, similar to my findings with the difference-in-differences approaches.  

<!-- 1-2 paragraphs contribution -->

This work makes several key contributions. First, this paper complements the literature on the economic impacts of invasive species [e.g., @epanchin-niell2017]. Recent work in economics has explored the impacts of the emerald ash borer [e.g., @jones_obesity_2023; @tan_save_2022], but to my knowledge, my work is the first to measure the ash borer's impact in a concentrated metropolitan region. It is also the first to examine invasive species' impacts on education outcomes. I also contribute to the literature detailing the environmental drivers of education performance, causally linking changes in vegetation to education outcomes for the first time outside of a laboratory setting. 

# Background

## Emerald ash borer in the Chicago region

One major threat to metropolitan and urban tree populations is the introduction of insect pests. The emerald ash borer is one such pest, first detected in North America in 2002. While it is one of several introduced pests that has affected US tree cover, it may represent a worst-case scenario [@herms2014]. The ash borer was first detected in North America in 2002 and has been referred to as the most destructive forest pest ever introduced to the United States [@nowak2016]. Ash borer exclusively target ash trees, and infestation is essentially fatal to any infested tree, so the arrival of the pest has meant the death or removal of millions of ash trees across the country. 

The Chicago region is the third-largest metropolitan region in the United States, and lies in a region heavily affected by emerald ash borer. The impacts on the region's ash populations has been well recognized. Prior to the arrival of the ash borer, ash trees were the most numerous non-invasive tree species in the region (Morton Arboretum 2020). However, a Chicago region tree census revealed that the area's standing ash population nearly halved between 2010 and 2020, dropping from an estimated 13 million to under 7 million (Morton Arboretum 2020). Of those 7 million remaining standing trees, 4 million are either dead or in decline. Further, as of 2020, more than 30\% of ash trees in the region are saplings, likely having regenerated from removed adults. Although many ash trees were replaced with alternative species and canopy cover broadly increased across the region, the overall number of large trees (> 6 inch diameter) dropped as mature ash were replaced by smaller trees. 

<!-- In 2013, healthy ash trees were treated with a preventative innoculation for the first time. This incesticide is injected into the trunk and prevents infestation in uninfested ash trees for up to three years. There has been previous work on the negative human impacts that may arise from the use of pesticides [@frank; @taylor].  -->

There exists previous work on impacts of ash borer and other forest attacking pests in the economics literature. For example, @tan_save_2022 uses staggered county-level ash borer detection to show its impact on mortality, through a channel of increased pollution. @jones_obesity_2023 use a similar approach to show that ash borer arrival led to reductions in actvity levels and increased rates of obesity in affected counties. @druckenmiller2020 instruments for tree mortality in the American West using the temperature threshold at which bark beetles experience winter die-off. They find that beetle-induced tree mortality decreased the value of timber tracts and home values and reduced hazard protection from air pollution, flood risk, and burn area. 

## Educational responses to ecosystem disturbance and vegetation

Trees provide substantial ecosystem services, particularly in metropolitan settings. These benefits include reduction of traffic pollution, psychological benefits, and moderation of hot temperatures. The association between the amount of trees and improved educational attainment and performance has been well documented. For example, @kweon2017 find that Washington DC schools with more trees had a higher percentage of proficient or advanced standardized test scores. However, despite vast amounts of correlational research, there exists scant causal evidence outside of the laboratory setting. Children exposed to more green vegetation show enhanced cognitive development and higher scores on cognitive development tests [@dadvand2015]. Green environments, such as open spaces with big trees, are even related to reduced symptoms of ADD and ADHD [@faber2009]. 

These improved outcomes may operate through several different channels. Urban tree cover is known to mitigate traffic-related air pollution [@nowak2006]. Trees may also reduce noise, allowing students to better focus [@gidlof-gunnarsson2007]. There are also psychological benefits associated with increased tree cover in neighborhoods and surrounding schools. @li2016 found that students who had views of trees and green environment from their classrooms, as compared to being in a room without windows or a room with a view of a brick wall, scored substantially higher on tests measuring attention. Studies consistently show a positive relationship between natural landscapes and enhanced physical activity amongst younger students [@dyment2007].

While causal studies detailing the effects of tree cover on educational outcomes are difficult to find, there exists a significant literature on the education impacts of ecosystem disturbances more broadly. High temperature has consistently been shown to affect learning [e.g., @park_hot_2022; @park2020]. For example, @park_heat_2020 show that hotter school days in the years leading up to the PSAT reduce scores, with extreme heat being particularly damaging. Pollution has also been shown to reduce performance on tests. @wen_smoke_2022 find that wildfire smoke exposure in the year leading up to a standardized test negatively affects performance. 

# Data and descriptive statistics {#data}

## Emerald ash borer survey

While infestations expand over relatively short distances through natural dispersal, infestations may go undetected for long periods. Long-distance spread occurs when infested material such as nursery stock or firewood is moved, potentially spurring new satellite populations. EAB infestation is fatal to all ash trees, however, it is difficult to detect until a tree is extensively damaged by the ash borer and begins to show symptoms. 

After the first ash borer detections in 2006, the Illinois Department of Agriculture (IDA) initiated survey efforts to determine the extent of ash borer spread. The IDA survey consisted of destructive bark peeling of selected trees. Selected trees were generally 4-8 diameters in width and in areas of easy and clear right-of-way access, with efforts to sample 1 tree per 4 square miles. The Chicago metropolitan region was the priority, but other parts of the state were surveyed as well, particularly if an infestation was suspected. Initially, damage was minimal as the detection method results were mostly negative, but positive finds became more and more prevalent. Ultimately, the state stopped survey efforts in 2015, as ash borer spread had become extensive. Figure \@ref(fig:survey) displays the locations of confirmed EAB infestations by year through the survey. 

```{r survey, echo=FALSE, fig.cap="Locations of confirmed ash borer infestations across Illinois based on IDA survey", out.width = '85%', fig.show='hold',fig.align='center'}

knitr::include_graphics(
  "figs/IL_eab_infestations.png"
                        )

```

## Chicago metropolitan area tree cover

I use two main data sources to understand how ash borer affected tree cover in the Metropolitan Chicago region. I first use annual 30m resolution tree loss and tree gain maps developed in @mccabe_prey_2018. These maps span 2000 to 2015 and importantly, allow me to differentiate changes in tree cover gain from tree cover loss. In order to better understand overall canopy cover trends and impacts, I also use annual 30m maps of canopy cover spanning from 1990 to 2017 developed in @hooper_spatial_2018 These yield the random forest probability that a pixel is canopy cover in a given year. 

The biggest concern with both of these data products is low resolution. Many ash trees in metropolitan areas are street trees, the loss of which may not manifest in a 30m resolution product. This fact makes the @hooper_spatial_2018 product particularly valuable, because while small canopy cover changes may not manifest in a binary canopy cover classification, they are likely reflected in the latent probability metric. Imposing a forest classification threshold would needlessly sacrifice information on possible canopy cover changes, giving this product an advantage over typical binary canopy cover classification products.  

- I have also obtained a 1m resolution map from Chicago's Morton Arboretum detailing canopy cover change between 2010 and 2018 (measured only in the two endpoint periods, not annual) but am currently unsure whether there is substantial benefit to integrating these data into the analysis. 

- @hooper_spatial_2018 also produce an annual product with NLCD classes, which distinguishes between deciduous and evergreen forest. Using evergreen as an outcome could be a good falsification test. 

## Education and test score data

Education data come from the Illinois State Board of Education (ISBE), which release annual reports on the performance of each school in the state on a number of metrics. I geocode locations of each public school in the state of Illinois using addresses provided by ISBE. This allows me to know the location of each school relative to confirmed ash borer infestations. I take advantage of two primary outcomes: 1) attendance rates; and 2) standardized test performance. 

The main standardized test I use is the Illinois Standards Achievement Test (ISAT). The ISAT was instituted for the purpose of identifying failing schools, and students were tested in reading and math from grades 3–8. ISBE reported composite school-level performance on the test between 2003 and 2014, when the ISAT was retired. These data include the percentage of students who meet or exceed standards set by the state for a given year. Many schools also separately report these statistics across low vs. non-low income students for each grade. 

ISBE also reports annual descriptive statistics about each school. For example, these data include low-income students, english-as-a-second language students, racial demographics, and enrollment. I use these as controls in my difference-in-differences estimation strategy. 

## Descriptive statistics

- I need to do more descriptive work

# Empirical strategy


## Difference-in-differences methods

Given the idiosynchratic nature of ash borer spread through insect flight or accidental transportation, the timing of confirmed detections can be thought of as quasi-random. These quasi-random infestation confirmations provide me with a shock to tree communities and canopy cover. I define treatment status using confirmed infestations from the IDA survey as described previously. An IDA confirmed infestation indicates that not only are trees in the vicinity infested and ultimately likely to die, but that community officials are aware of the need for tree removal and replacement. Because infestations are often hard to detect and trees typically do not die from infestation immediately, a confirmed infestation can be thought of as an exogenous event that spurs new removal and replacement activity.

I use difference-in-differences (DID) approaches to evaluate the contemporaneous and dynamic effects of ash borer infestation. Identification relies on a common trends assumption, which amounts to assuming that outcomes in the treatment group would have followed the same evolution as those in the control and not yet treated groups, had treatment never occured. 

Limitations to the DID approach arise from the distinction between confirmed ash borer infestation and ash borer presence. It is possible, and perhaps even likely, that ash borer are present in the area for some amount of time prior to confirmed infestation. Further, nearby infestations may go unreported if survey efforts are stopped near already confirmed infested areas. If this were the case, some infested areas may be incorrectly classified as untreated. In the context of my DID approaches, this should underestimate the impacts of the ash borer.

### Tree cover impacts

I begin by examining the impact of ash borer infestation on canopy cover outcomes. This establishes the pathway through which ash borer detection is likely to affect education outcomes. In other words, hypothesized mechanisms by which we expect infestations to affect education outcomes operate through a change in canopy cover. The unit of analysis is the grid cell, $i$, where my preferred grid cell size is $5\times 5$ km. I also show results based on 3km grid cells in the appendix. My first specification is a two-way fixed effects regression of the following form:

\begin{align}
tree_{it}  = & \beta \times infestation_{it} + \gamma_i + \lambda_t + X_{it} + u_{it}
\end{align}

where:

- $tree_{it}$ represents the tree cover outcome within grid $i$ in time $t$. The three tree cover outcomes are tree cover loss, tree cover gain, and mean grid cover canopy cover probability as described in section \@ref(data) 
- $\gamma_i$ and $\lambda_t$ represent grid and year fixed effects, respectively

In addition to regressions of the form above, I use the DID estimator developed in @callaway2020. It is robust to general treatment effect heterogeneity that may plague two-way fixed effects regressions [@de_chaisemartin2020]. Causal interpretation also relies on a conditional common trends assumption, meaning that common trends need hold only after conditioning on pre-treatment covariates. All event study plots presented in this paper are based on estimates from the @callaway2020 estimator. 

### Education impacts 

I examine the impact of ash borer infestation on school-level outcomes using the same DID strategies as above. The unit of analysis is the individual school, $j$, and school $j$ is considered treated in the years after an infestation is detected within a 2.5 km radius of the school. The two-way fixed effects regression is now as follows:

\begin{align}
education_{jt}  = & \beta \times infestation_{jt} + \gamma_j + \lambda_t + X_{jt} + u_{jt}
\end{align}

- $education_{jt}$ represents the education outcome of interest for school $j$ in time $t$

Again, I also estimate these treatment effects with the @callaway2020 estimator. 

## Dose response of education outcomes to canopy cover change

Of particular interest is to determine the dose response of education outcomes to canopy cover directly. I use an instrumental variables approach to understand the dose response of education outcomes to canopy cover. This involves instrumenting for canopy cover change with ash borer infestation. The first stage of the 2SLS is:

$$ treechange_{jt} = \psi \times infestation_{jt} + \gamma_j + \lambda_t + X_{jt} + u_{jt}$$

and the second stage is then,

$$ education_{jt} =  \alpha \times \widehat{treechange}_{jt} + \gamma_j + \lambda_t + X_{jt} + e_{jt}$$


- $\widehat{treechange}_{jt}$ represents the rate of tree cover change within the 2.5 km surrounding school $j$ in time $t$

- $education_{jt}$ represents the various education outcomes of interest

The causal interpretation of $\alpha$ relies on an exclusion restriction, which in this context, requires that confirmed ash borer infestation only affects education outcomes through the channel of canopy cover change. Here, I cannot separately examine the effects of tree cover gain and loss without violating the exclusion restriction by construction. 

There are potential threats to the plausibility of the exclusion restriction in this context. In particular, I worry that the use of pesticides may violate this restriction. Pesticides have been shown to negatively impact health outcomes [e.g., @taylor_cicadian_2022], and they have been used to combat the ash borer. In 2013, healthy ash trees were treated with a preventative innoculation for the first time. This incesticide is injected into the trunk and prevents infestation in uninfested ash trees for up to three years. If infestation led to increased pesticide use, which in turn led to adverse education outcomes, this could violate the exclusion restriction. 

I was leaning against including the IV in this paper, but maybe I will subset the IV analysis to pre-2013 years as a robustness check in future iterations of the paper. I don't show the results from the IV at this time. 

# Main results

## Tree cover impacts

In this section, I aim to establish that confirmed EAB infestations do indeed have significant canopy cover effects. There are two main channels through which ash borer detection may result in tree loss. The first is through the EAB directly, as an ash tree will die between one and four years following infestation, depending on the size and health of the individual ash tree. The second is through intentional removal of infested trees. Most communities declare any confirmed infested tree a public nuisance and require that the tree be removed [e.g., @macomb2007]. As such, a confirmed infestation is likely to lead to manual removal of infested or dead trees in the vicinity.

Because removed trees are likely replaced with new trees, some may question why we should observe canopy cover loss on an annual scale. Most replacement trees are unlikely to generate the canopy cover of large healthy trees. A newly planted tree may have trunk diameter as small as two inches and provide little to no canopy cover, while a large healthy ash may grow 60 feet tall and 25 to 40 feet wide (Morton Arboretum). 

Ash borer detection may also lead to lower levels of tree cover gain. Removal of damaged or dead trees is costly to individuals and communities. Estimates suggest a cost near \$1,000 to remove and replant a single tree. More broadly, EAB had a massive impact on forestry budgets across the United States [@hauer2017]. While budgets in states with confirmed EAB infestation saw sizable increases in tree removal budgets relative to non-EAB confirmed states, budgets for tree planting did not change. Because most removed trees are likely replaced, it is plausible that trees that would've been planted in the absence of EAB were never established. In other words, funding allocated for establishing new tree cover instead went toward replacement of EAB infested trees. 

Table 1 shows treatment effect estimates based on equation 1. Canopy cover is reduced by roughly 2\% following ash borer infestation detection, as indicated by the first two columns. We also see that there is both a decline in rates of tree cover gain and an increase in rates of tree cover loss, although the significance of these estimates varies across the two estimators. Both prior to and after the arrival of the ash borer to the Chicago area, tree cover gain outpaced loss across the study area. Nearly identical results based on 3 kilometer grid cells can be seen in section \@ref(#grid-3km).

```{r results-5km}
main_results_5km <- readRDS("results/main_results_5km.rds")%>%
  mutate(crit = ATT/se)%>%
  mutate_at(vars(ATT, se), ~ round(., digits = 4))

twfe_results <- main_results_5km %>% filter(estimator == "TWFE")
  
cs_results <- main_results_5km %>% filter(estimator == "CS")

```


\begin{table}[H]\centering
\caption{Estimated ATT of EAB infestation on tree cover metrics}
\smallskip
\begin{tabular}{l c c c c c c}

\toprule
 & \multicolumn{2}{c}{Canopy cover} & \multicolumn{2}{c}{Tree loss (acres/year)} & \multicolumn{2}{c}{Tree gain (acres/year)} \\
 
 \cmidrule(lr){2-3} \cmidrule(lr){4-5}  \cmidrule(lr){6-7}
 
  Estimator: & TWFE & C\&S & TWFE & C\&S & TWFE & C\&S\\ 
\midrule
 $\widehat{ATT}$  & `r twfe_results$ATT[1]`***  & `r cs_results$ATT[1]`***  & `r twfe_results$ATT[2]`***  & `r cs_results$ATT[2]`& `r twfe_results$ATT[3]`***  & `r cs_results$ATT[3]`   \\
                    & (`r twfe_results$se[1]`) & (`r cs_results$se[1]`) & (`r twfe_results$se[2]`) & (`r cs_results$se[2]`) & (`r twfe_results$se[3]`) & (`r cs_results$se[3]`) \\ 
\midrule
Pre-treatment mean  & `r twfe_results$pre.treat[1]`  & `r cs_results$pre.treat[1]`  & `r twfe_results$pre.treat[2]`  & `r cs_results$pre.treat[2]`& `r twfe_results$pre.treat[3]`  & `r cs_results$pre.treat[3]`   \\
Grid cell area      &     $25km^2$    &  $25km^2$ &     $25km^2$    &  $25km^2$ &     $25km^2$    &  $25km^2$  \\ 
 Number of grid cells      & `r main_results_5km$Ngrid[1]` & `r main_results_5km$Ngrid[1]` & `r main_results_5km$Ngrid[1]` & `r main_results_5km$Ngrid[1]` & `r main_results_5km$Ngrid[1]` & `r main_results_5km$Ngrid[1]`  \\          
\bottomrule
\addlinespace[1ex]
\multicolumn{3}{l}{
  \textsuperscript{*}$p<0.1$,
  \textsuperscript{**}$p<0.05$, 
  \textsuperscript{***}$p<0.01$
}
\end{tabular}
\end{table}

Figure \@ref(fig:canopy-grid-es) displays event study estimates in order to understand the dynamics of canopy cover change following infestation. This figure also allows us to gauge the plausibility of the common trends assumption on which this identification strategy relies. The assumption does seem plausible given similar pre-trends in outcomes as well as the quasi-random nature of EAB infestation confirmation timing. However, there appears to be a slight dip in canopy cover leading up to the confirmation year. This is likely because ash borer are present prior to the actual infestation confirmation date. 

```{r canopy-grid-es, echo=FALSE, fig.cap="Canopy cover impacts of emerald ash borer infestation through time. This plot shows treatment effect estimates from the Callaway annd Sant'anna (2020) estimator by event time. Canopy cover probability declines in infested areas following the year in which an infestation is first confirmed (e = 0).", out.width = '85%', fig.show='hold',fig.align='center'}

knitr::include_graphics(
  "results/canopy_es_grid5km.png"
                        )

```

\clearpage

## Education results

### Main outcomes

I now aim to show evidence of the ash borer's impact on education outcomes. Table 2 shows how ash borer infestation impacted my education outcomes of interest. Infestation leads to a roughly 2\% reduction in the percentage of students meeting state standards on the ISAT. The same can be said for standardized tests more broadly, as evidenced by the second set of estimates. There is some weak evidence that attendance rates are affected, however, the significance is variable depending on the estimator. Further, the magnitude of the effect is less than a 1\% reduction, suggesting that attendance rates are not as concerning as the declines in test score performance. 

```{r school-results-educ}

twfe_results <- readRDS("results/school_twfe_results_3km.rds")[4:7,]%>%
  mutate(crit = ATT/se)%>%
  mutate_at(vars(ATT, se), ~ round(., digits = 4))
  
cs_results <- readRDS("results/school_cs_results_3km.rds")[4:7,]%>%
  mutate(crit = ATT/se)%>%
  mutate_at(vars(ATT, se), ~ round(., digits = 4))

```

\begin{table}[H]\centering
\caption{estimated ATT of EAB infestation on education outcomes}
\smallskip
\begin{tabular}{l c c c c c c}

\toprule
Outcome: & \multicolumn{2}{c}{ISAT composite} & \multicolumn{2}{c}{All tests} &  \multicolumn{2}{c}{Attendance rate} \\
 
 \cmidrule(lr){2-3} \cmidrule(lr){4-5}  \cmidrule(lr){6-7}
 
  Estimator: & TWFE & C\&S & TWFE & C\&S & TWFE & C\&S\\ 
\midrule
 $\widehat{ATT}$  & `r twfe_results$ATT[1]`***  & `r cs_results$ATT[1]`***  & `r twfe_results$ATT[2]`***  & `r cs_results$ATT[2]`*** & `r twfe_results$ATT[3]`**  & `r cs_results$ATT[3]` \\
                    & `r twfe_results$se[1]`  & `r cs_results$se[1]`  & `r twfe_results$se[2]`  & `r cs_results$se[2]` & `r twfe_results$se[3]`  & `r cs_results$se[3]` \\
\midrule
Pre-treatment mean   & `r twfe_results$pre.treat[1]`  & `r twfe_results$pre.treat[1]`  & `r twfe_results$pre.treat[2]`  & `r twfe_results$pre.treat[2]` & `r twfe_results$pre.treat[3]`  & `r twfe_results$pre.treat[3]` \\ 
 Number of schools      &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]`   \\          
\bottomrule
\addlinespace[1ex]
\multicolumn{3}{l}{
  \textsuperscript{*}$p<0.1$,
  \textsuperscript{**}$p<0.05$, 
  \textsuperscript{***}$p<0.01$
}
\end{tabular}
\end{table}

I present event study estimates of the impact of ash borer infestation in Figure \@ref(fig:isat-es). We see that test performance steadily declines following ash borer infestation detection. Looking at the treatment effect estimates prior to treatment, which can be interpreted as pseudo-ATTs, there is some some evidence of stable pre-trends. Despite the appearance of a slight increasing trend prior to infestation detection, the effect reverses this trend. A more concerning observation would be if the trend were already sloping downwards in advance of the treatment date. 

```{r isat-es, echo=FALSE, fig.cap="ISAT impacts of emerald ash borer infestation through time. This plot shows treatment effect estimates from the Callaway annd Sant'anna (2020) estimator by event time. The outcome in this plot is the composite percentage of students meeting or exceeding the achievement threshold set by the state of Illinois in a given year.", out.width = '85%', fig.show='hold',fig.align='center'}

knitr::include_graphics(
  "results/isat_es_3km.png"
                        )

```
\clearpage

### ISAT achievement stratified by income and grade level

ISBE reports ISAT performance stratified by grade, subject, and income across four benchmarks: 1) academic warning, 2) below standard, 3) meets standard, and 4) exceeds standard. Figure \@ref(fig:isat-stratify) presents estimates on the impact of infestation across grade and income level for the Math portion of the ISAT. The left panel shows that infestation caused more low-income students to fall into the academic warning and below standard categories. These estimates suggest that low-income students are affected most by ash borer infestation. 

```{r isat-stratify, echo=FALSE, fig.cap="Left: This panel shows how infestation affected the percentage of students whose performance fell into academic warning based on their ISAT score.; Right: This panel shows how infestation affected the percentage of students whose performance did not meet the ISAT score threshold set by the state. This excludes students who fell into the academic warning category.", out.width = '49%', fig.show='hold',fig.align='center'}

knitr::include_graphics(
  c("results/schart_math_warning.jpeg", "results/schart_math_below.jpeg")
                        )

```

<!-- ## Dose response results using instrumental variables  -->

```{r iv-results}

iv_results <- readRDS("results/iv_results.rds")

```

<!-- \begin{table}[H]\centering -->
<!-- \caption{IV estimates of the impact of education responses to tree cover change} -->
<!-- \smallskip -->
<!-- \begin{tabular}{c c c c} -->
<!-- \toprule -->
<!--   Outcome: & ISAT composite & All tests & Attendance rates \\  -->
<!-- \midrule -->
<!--  $\widehat{\alpha}$  &   `r iv_results$coef[2]`***   & `r iv_results$coef[3]`*** & `r iv_results$coef[4]` \\ -->
<!--              &      (`r iv_results$se[2]`)   & (`r iv_results$se[3]`) & (`r iv_results$se[4]`)\\ -->
<!-- \midrule -->
<!-- First stage F-statistic   & `r iv_results$Fstat[2]`   & `r iv_results$Fstat[3]` & `r iv_results$Fstat[4]` \\  \\ -->
<!-- \bottomrule -->
<!-- \multicolumn{3}{l}{ -->
<!--   \textsuperscript{*}$p<0.1$, -->
<!--   \textsuperscript{**}$p<0.05$,  -->
<!--   \textsuperscript{***}$p<0.01$ -->
<!-- } -->
<!-- \end{tabular} -->
<!-- \end{table} -->

## Heterogeneity

- Effect is actually through ash trees
- Wealthier areas and even areas with more canopy cover see lower impacts from infestation

\input{results/twfe_het_5km.tex}

\clearpage

# Mechanisms {#mechanisms}

I've struggled a bit to successfully get at the expected mechanisms here. It is likely that it is a combination of mechanisms. I'll outline some of the mechanisms it would be nice to include and the progress (or lack thereof) I've made toward addressing each. 

## Pollution

I actually have had a bit of success here. There are several products that could work here. A quick pass seemed to indicate that pm2.5 increased in infestation areas after the confirmed infestation, but I need to be a bit more thorough before I add in. Will do this in the next week. 

## Temperature

My main struggle on the temperature front is resolution of the products I've explores thus far. For example, the resolution of the PRISM data is 4km, which is much too coarse to see much variation at the scale I'm working with. My next step will be to look for monitors but any advice or known data here would be great. 

I initially though I might be able to argue away temperature as a driver, since these tests are taken when temperatures are quite moderate in Chicago. However, much of the temperature and learning literature shows that extreme heat in the year leading up to the test can have substantial impact, even if not on test day. 

## Pesticide use

I've been able to obtain the location of treated ash trees within the city limits of Chicago. Unfortunately, the date is missing on most of these. Since, pesticides weren't used in the area until 2013, I will rerun some of these analyses in the years prior to that to provide confidence that pesticides do not drive the entire result. 

# Discussion

- will flesh out later

# References
<div id="refs"></div>

\renewcommand{\thesection}{A}


# Appendix

## Tree cover impacts based on 3 km grid cells {#grid-3km}

```{r results-3km}
main_results_3km <- readRDS("results/main_results_3km.rds")%>%
  mutate(crit = ATT/se)%>%
  mutate_at(vars(ATT, se), ~ round(., digits = 4))

twfe_results <- main_results_3km %>% filter(estimator == "TWFE")
  
cs_results <- main_results_3km %>% filter(estimator == "CS")

```


\begin{table}[H]\centering
\caption{Estimated ATT of EAB infestation on tree cover metrics}
\smallskip
\begin{tabular}{l c c c c c c}

\toprule
 & \multicolumn{2}{c}{Canopy cover} & \multicolumn{2}{c}{Tree loss (acres/year)} & \multicolumn{2}{c}{Tree gain (acres/year)} \\
 
 \cmidrule(lr){2-3} \cmidrule(lr){4-5}  \cmidrule(lr){6-7}
 
  Estimator: & TWFE & C\&S & TWFE & C\&S & TWFE & C\&S\\ 
\midrule
 $\widehat{ATT}$  & `r twfe_results$ATT[1]`***  & `r cs_results$ATT[1]`***  & `r twfe_results$ATT[2]`***  & `r cs_results$ATT[2]`& `r twfe_results$ATT[3]`***  & `r cs_results$ATT[3]`   \\
                    & (`r twfe_results$se[1]`) & (`r cs_results$se[1]`) & (`r twfe_results$se[2]`) & (`r cs_results$se[2]`) & (`r twfe_results$se[3]`) & (`r cs_results$se[3]`) \\ 
\midrule
Pre-treatment mean  & `r twfe_results$pre.treat[1]`  & `r cs_results$pre.treat[1]`  & `r twfe_results$pre.treat[2]`  & `r cs_results$pre.treat[2]`& `r twfe_results$pre.treat[3]`  & `r cs_results$pre.treat[3]`   \\
Grid cell area      &     9$km^2$    &  9$km^2$ &     9$km^2$    &  9$km^2$ &     9$km^2$   &  9$km^2$  \\ 
 Number of grid cells      & `r main_results_3km$Ngrid[1]` & `r main_results_3km$Ngrid[1]` & `r main_results_3km$Ngrid[1]` & `r main_results_3km$Ngrid[1]` & `r main_results_3km$Ngrid[1]` & `r main_results_3km$Ngrid[1]`  \\          
\bottomrule
\addlinespace[1ex]
\multicolumn{3}{l}{
  \textsuperscript{*}$p<0.1$,
  \textsuperscript{**}$p<0.05$, 
  \textsuperscript{***}$p<0.01$
}
\end{tabular}
\end{table}

\input{results/twfe_het_3km.tex}

## Ash borer impacts on tree cover in 2.5 km buffer surrounding schools {#school-tree}

```{r school-results-tree}

twfe_results <- readRDS("results/school_twfe_results_3km.rds")[1:3,]%>%
  mutate(crit = ATT/se)%>%
  mutate_at(vars(ATT, se), ~ round(., digits = 4))
  
cs_results <- readRDS("results/school_cs_results_3km.rds")[1:3,]%>%
  mutate(crit = ATT/se)%>%
  mutate_at(vars(ATT, se), ~ round(., digits = 4))

```

\begin{table}[H]\centering
\caption{Estimated ATT of EAB infestation on tree cover in school vicinity}
\smallskip
\begin{tabular}{l c c c c c c}

\toprule
 & \multicolumn{2}{c}{Canopy cover} & \multicolumn{2}{c}{Tree loss (acres/year)} & \multicolumn{2}{c}{Tree gain (acres/year)} \\
 
 \cmidrule(lr){2-3} \cmidrule(lr){4-5}  \cmidrule(lr){6-7}
 
  Estimator: & TWFE & C\&S & TWFE & C\&S & TWFE & C\&S\\ 
\midrule
 $\widehat{ATT}$  & `r twfe_results$ATT[1]`***  & `r cs_results$ATT[1]`***  & `r twfe_results$ATT[2]`  & `r cs_results$ATT[2]` & `r twfe_results$ATT[3]`***  & `r cs_results$ATT[3]`*** \\
                    & `r twfe_results$se[1]`  & `r cs_results$se[1]`  & `r twfe_results$se[2]`  & `r cs_results$se[2]` & `r twfe_results$se[3]`  & `r cs_results$se[3]` \\
\midrule
 Number of schools      &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]` &   `r twfe_results$Nschools[1]`   \\          
\bottomrule
\addlinespace[1ex]
\multicolumn{3}{l}{
  \textsuperscript{*}$p<0.1$,
  \textsuperscript{**}$p<0.05$, 
  \textsuperscript{***}$p<0.01$
}
\end{tabular}
\end{table}
